Formalization of Forcing in Isabelle/ZF

### 2020-06-10 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
first draft of some structure for slides

\usepackage[utf8]{inputenc}
%%%
%%% Natbib must be here!! (after Beamer stuff)
\usepackage[numbers]{natbib}
\input{header-slides}
\usepackage{isabelle,isabellesym}

@@

\begin{document}
\begin{frame}[plain]
  \titlepage
  \begin{center}
    \insertlogo
  \end{center}
\end{frame}
%
\begin{frame}
  \frametitle{Summary}
  \tableofcontents
  \transwipe
\end{frame}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{}

@@

@@

>>>>

\usepackage[utf8]{inputenc}
%%%
%%% Natbib must be here!! (after Beamer stuff)
\usepackage[authoryear,square]{natbib}
\input{header-slides}
\usepackage{isabelle,isabellesym}
\newcommand{\uscore}{\isacharunderscore}

@@

\begin{document}
%\nobibliography*

\begin{frame}[plain]
  \titlepage
  \begin{center}
    \insertlogo
  \end{center}
\end{frame}
%
\begin{frame}
  \frametitle{Summary}
  \tableofcontents
  \transwipe
\end{frame}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Why Isabelle/ZF?}

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
\end{frame}

\begin{frame}
  \frametitle{}

\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{Paulson's \texttt{ZF-Constructible}}

\begin{frame}
  \frametitle{Paulson's \texttt{ZF-Constructible}}
\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{The ctm approach to forcing}

\begin{frame}
  \frametitle{The ctm approach to forcing, I}
\end{frame}

\begin{frame}
  \frametitle{The ctm approach to forcing, II}
\end{frame}

\begin{frame}
  \frametitle{Other approaches}
  \begin{itemize}
  \item<+-> \emph{Set theory in HOL}
    \begin{itemize}
    \item \isatt{HOLZF} \citep{DBLP:conf/ictac/Obua06}
    \item \isatt{ZFC{\uscore}in{\uscore}HOL}
      \citep{ZFC_in_HOL-AFP}
    \end{itemize}
  \item<+-> \alert{Lean}: Full formalization of the Boolean-valued approach to
    forcing and the independence of $\CH$ \citep{DBLP:conf/cpp/HanD20}.
  \end{itemize}
  %
  \begin{block}{A word on consistency}<+->
    \begin{description}
      \setlength{\labelwidth}{12em}
      \setlength{\labelsep}{2em}
      \setlength{\itemindent}{6em}
    \item[Isabelle/ZF + ctm]  (far) less than $\ZF$ +  one
      inaccessible.
    \item[\isatt{HOLZF},
      \isatt{ZFC{\uscore}in{\uscore}HOL}]
      approximately $\ZF$ +  one   inaccessible. 
    \item[Lean (CiC)] $\ZF$ + $\omega$ inaccessibles. 
    \end{description}
  \end{block}
\end{frame}

%% \begin{frame}
%%   \frametitle{A word on consistency}
%% \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The development}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{What did we accomplished?}

\begin{frame}
  \frametitle{What did we accomplished?}
  %
  \begin{shadowblock}{}
    \begin{enumerate}
    \item We adapted \texttt{ZF-Constructible} to obtain
      absoluteness results for nonempty transitive classes ($\leftarrow$
      already in the 2020 release of Isabelle).
    \item We formalized the forcing relation and proved the Fundamental
      Theorems.
    \item We showed that generic extensions of ctms of $\ZF$ are also
      ctms of $\ZF$ (respectively, adding $\AC$).
    \item We provided the forcing notion that adds a Cohen real,
      therefore proving the existence of a nontrivial extension.
    \end{enumerate}
  \end{shadowblock}
\end{frame}

%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{A map}

\begin{frame}
  \frametitle{A map to the development}
\end{frame}

\begin{frame}
  \frametitle{Locale structure} 
             {  
               \renewcommand{\arraystretch}{1.5}               \begin{tabular}{rcl}
                 \texttt{forcing{\uscore}notion} & = & preorder $\PP$ with top. \\
                 %    \texttt{M{\uscore}ZF} & = & set model of the $\ZF$ axioms \\ 
                 \texttt{M{\uscore}ZF{\uscore}trans} & = & set model $M$ of the $\ZF$
                 axioms \alert{+}  $M$ transitive \\ 
                 \texttt{M{\uscore}ctm} & = &  \texttt{M{\uscore}ZF{\uscore}trans} \alert{+}
                 $M$ countable \\
                 \texttt{forcing{\uscore}data} & =  & \texttt{M{\uscore}ctm} \alert{+}
                 \texttt{forcing{\uscore}notion} in $M$\\
                 \texttt{separative{\uscore}notion} & = &
                 \texttt{forcing{\uscore}notion} \alert{+} $\PP$ separative \\
                 \texttt{M{\uscore}ctm{\uscore}separative} & = &
                 \texttt{forcing{\uscore}data} \alert{+} \texttt{separative{\uscore}notion}
               \end{tabular} 
             }
\end{frame}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{}
%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
\subsection{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Looking forward}

\begin{frame}
  \frametitle{Looking forward}
  \begin{block}{Formalizing math}
    \begin{itemize}
    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
    \item<+-> Theorems on preservation of cardinals.
    \end{itemize}
  \end{block}
  %
  \begin{block}{Technical aids}<+->
    \begin{enumerate}
    \item<.-> Automatic relativization of concepts.
    \item<+-> Relative functions.
    \item<+->
    \end{enumerate}
  \end{block}
\end{frame}

@@

\begin{frame}
  \frametitle{References}
  \bibliographystyle{mi-estilo-else}
  \bibliography{../forcing_in_isabelle_zf}
\end{frame}
